MS Technical Paper: \\ Placement Algorithms for Heterogeneous FPGAs

### 2025-05-07 latex/paper/sections/09-packing.tex
netlist stuff

\documentclass{article}
\usepackage{hyperref}
\usepackage{enumitem}
\usepackage{graphicx}
\usepackage{amsmath}
\usepackage{mathpazo}
\usepackage{xcolor}
\usepackage{float}
\usepackage{listings}
\lstset{
  basicstyle=\ttfamily\small,
  breaklines=true,
  columns=fullflexible,
  frame=single,
  showspaces=false,
  showtabs=false,
  keepspaces=false
}
\usepackage[export]{adjustbox}
\usepackage{multicol}
\usepackage{caption}
\usepackage[a4paper,
            bindingoffset=0.2in,
            left=0.5in,
            right=0.5in,
            top=0.5in,
            bottom=0.5in,
            footskip=.25in]{geometry}

\begin{document}
\title{MS Technical Paper: \\ Placement Algorithms for Heterogeneous FPGAs}
\author{Brian B Cheng \\ Rutgers University Department of Electrical and Computer Engineering}

\date{}
\maketitle

\begin{multicols}{2}

\section{Keywords}
\begin{itemize}
    \item FPGA, EDA, Placement, Simulated Annealing, Optimization, RapidWright
\end{itemize}

\section{Abstract}
    fdsafdsafdsa. \\ 
    fdsafdsafdsa. \\ 
    fdsafdsafdsa. \\ 
    fdsafdsafdsa. \\ 
    fdsafdsafdsa. \\ 
    fdsafdsafdsa. \\ 
    fdsafdsafdsa. \\ 
    fdsafdsafdsa. \\ 
    fdsafdsafdsa. \\ 
    fdsafdsafdsa. \\ 
    fdsafdsafdsa. \\ 
    fdsafdsafdsa. \\ 
    fdsafdsafdsa. \\

\section{Introduction}

    Field-Programmable Gate Arrays (FPGAs) have witnessed rapid growth in capacity and versatility, driving significant advances in computer-aided design (CAD) and electronic design automation (EDA) methodologies. 
    Since the early-to-mid 2000s, the stagnation of single-processor performance relative to the rapid increase in integrated circuit sizes has led to a design productivity gap, where the computational effort for designing complex chips continues to rise. 
    FPGA CAD flows mainly encompass synthesis, placement, and routing; all of which are NP-hard problems, of which placement is one of the most time-consuming processes. 
    Inefficienct placement strategy not only extends design times from hours to days, thereby elevating cost and reducing engineering productivity, but also limits the broader adoption of FPGAs by software engineers who expect compile times akin to those of conventional software compilers like {\tt gcc}.

    For these reasons, FPGA placement remains a critical research effort even today. 
    In this paper, we study and implement established placement methods. 
    To do this, we use the RapidWright API, which is a semi-open-source research effort from AMD/Xilinx that enables custom solutions to FPGA design implementations and design tools that are not offered by their industry-standard FPGA environment, Vivado. 
    We implement multiple variations of simulated annealing placers for Xilinx's 7-series FPGAs, with an emphasis on minimizing total wirelength while mitigating runtime. 
    Our implementation is organized into three consecutive substages. 
    The \textbf{prepacking} stage involves traversing a raw EDIF netlist to identify recurring cell patterns—such as CARRY chains, DSP cascades, and LUT-FF pairs—that are critical for efficient mapping and legalization. 
    In the subsequent \textbf{packing} stage, these identified patterns, along with any remaining loose cells, are consolidated into SiteInst objects that encapsulate the FPGA’s discrete resource constraints and architectural nuances. 
    Finally, the \textbf{placement} stage employs a simulated annealing (SA) algorithm to optimally assign SiteInst objects to physical sites, aiming to minimize total wirelength while adhering to the constraints of the 7-series architecture.

    Simulated annealing iteratively swaps placement objects guided by a cost function that decides which swaps should be accepted or rejected. 
    Hill climbing is permitted by occasionally accepting moves that increase cost, in hope that such swaps may later lead to a better final solution. 
    SA remains a popular approach in FPGA placement research due to its simplicity and robustness in handling the discrete architectural constraints of FPGA devices. 
    While SA yields surprisingly good results given relatively simple rules, it is ultimately a heuristic and stochastic approach that explores the vast placement space by making random moves. 
    Most of these moves will be rejected, meaning that SA must run many iterations, usually hundreds to thousands, to arrive at a desirable solution.

    In the ASIC domain, where placers must handle designs with millions of cells, the SA approach has largely been abandoned in favor of analytical techniques, owing to SA's runtime and poor scalability. 
    Modern FPGA placers have also followed suit, as new legalization strategies allow FPGA placers to leverage traditionally ASIC placement algorithms and adapt them to the discrete constraints of FPGA architectures. 
    While this paper does not present a working analytical placer, it will explore ways to build upon our existing infrastructure (prepacker and packer) to replace SA with AP.

    The paper first begins by elaborating on general FPGA architecture and then specifically the Xilinx 7-Series architecture. 
    Then, the paper will elaborate on the FPGA design flow, then the role that the RapidWright API plays in the design flow. 
    We explain in detail each of these concepts for a broader audience as they provide much needed context for FPGA placement algorithms as a concept. 
    However, readers who are already familiar with these concepts can skip directly to the RapidWright API section \ref{sec:rapidwright_api} or to the Simulated Annealing section \ref{sec:simulated_annealing}.

\section{FPGA Architecture History}

    Before any work can begin on an FPGA placer, it is necessary to understand both the objects being placed and the medium in which they are placed.
    Configurable logic devices have undergone significant evolution over the past four decades. 
    We will briefly review the evolution of configurable logic architecture starting in the 1970s and quickly work our way up to modern day FPGA architecture.

    \textbf{PLA:} \quad
    The journey began with the Programmable Logic Array (PLA) in the early 1970s. 
    The PLA implemented output logic using a programmable-OR and programmable-AND plane that formed a sum-of-products equation for each output through programmable fuses. 
    Around the same time, the Programmable Array Logic (PAL) was introduced. 
    The PAL simplified the PLA by fixing the OR gates, resulting in a fixed-OR, programmable-AND design, which sacrificed some logic flexibility to simplify its manufacture. 
    Figure \ref{fig:pal_2} shows one such PAL architecture.

    \textbf{CPLD:} \quad 
    Later in the same decade came the Complex Programmable Logic Device (CPLD), which took the form of an array of Configurable Logic Blocks (CLBs). 
    These CLBs were typically modified PAL blocks that included the PAL itself along with macrocells such as flip-flops, multiplexers, and tri-state buffers. 
    The CPLD functioned as an array of PALs connected by a central programmable switch matrix and could be programmed using a hardware description language (HDL) like VHDL. 
    Figure \ref{fig:cpld} shows one such CPLD architecture. 
    \\

    {
        \centering
        \includegraphics[width=\columnwidth]{figures/pal_2.png}
        \captionof{figure}{PAL architecture with 5 inputs, 8 programmable AND gates and 4 fixed OR gates}
        \label{fig:pal_2}
        % https://www.electronics-tutorial.net/Programmable-Logic-Device-Architectures/Programmable-Logic-Devices/Programmable-Array-Logic-PAL/
        % https://www.naukri.com/code360/library/difference-between-pla-and-pal 
    }

    \newcolumn

    \textbf{Homogeneous FPGA:} \quad 
    The mid-1980s saw the introduction of homogeneous FPGAs, which were built as a grid of CLBs. 
    Rather than using a central programmable switch matrix as in CPLDs, FPGAs adopted an island style architecture in which each CLB is surrounded on all sides by programmable routing resources, as shown in Figure \ref{fig:fpga}. 
    The first commercially viable FPGA, produced by Xilinx in 1984, featured 16 CLBs arranged in a 4x4 grid. 
    As FPGA technology advanced, CLBs were redesigned to use lookup tables (LUTs) instead of PAL arrays for greater logic density. 
    The capacity of an FPGA was often measured by how many logical elements or CLBs it offered, which grew from hundreds to thousands and now to hundreds of thousands of CLBs.
    \\

    {
        \centering
        \includegraphics[width=0.8\columnwidth]{figures/cpld.png}
        \captionof{figure}{CPLD architecture with 4 CLBs (PAL-like blocks)}
        \label{fig:cpld}
        % https://www.electronics-tutorial.net/Programmable-Logic-Device-Architectures/CPLD/Complex-Programmable-Logic-Device-CPLDs/
    }
    {
        \centering
        \includegraphics[width=\columnwidth]{figures/homogenous_fpga.png}
        \captionof{figure}{A homogeneous island-style FPGA architecture with 16 CLBs in a grid.}
        \label{fig:fpga}
        % https://www.electronics-tutorial.net/Programmable-Logic-Device-Architectures/FPGA/Field-programmable-gate-array-FPGA/
    }

    \newpage

    {
        \centering
        \includegraphics[width=\columnwidth]{figures/heterogenous_fpga_3.jpg}
        \captionof{figure}{A heterogeneous island-style FPGA with a mix of CLBs and macrocells.}
        \label{fig:fpga}
        % https://www.electronics-tutorial.net/Programmable-Logic-Device-Architectures/FPGA/Field-programmable-gate-array-FPGA/
    }
    \vspace{0.5cm}

    \textbf{Heterogenous FPGA:} \quad 
    This brings us to modern day FPGA architectures. 
    To meet the needs of increasingly complex designs, FPGA vendors introduced heterogeneous FPGAs. 
    In these devices, hard macros such as Block RAM (BRAM) and Digital Signal Processing (DSP) slices are integrated into the programmable logic fabric along with CLBs, like shown in Figure \ref{fig:fpga}. 
    This design enables the direct instantiation of common subsystems like memories and multipliers, without having to recreate them from scratch using CLBs. 
    Major vendors such as Xilinx and Altera now employ heterogeneous island-style architectures in their devices. 
    As designs become increasingly large and complex, FPGAs meet the demand by becoming increasingly heterogenous, incorporating a wider variety of hard macros into the fabric.

\section{Xilinx 7-Series Architecture}
    \label{sec:7_series}
    The Xilinx 7-Series devices, first introduced in 2010, follow a heterogeneous island-style architecture as discussed previously. 
    Although the 7-Series was later superseded in 2013 by the UltraScale architecture, the 7-Series remains highly relevant due to its accessibility, wide availability, and compatibility with open-source tooling. 
    Representative sub-families include Artix-7, Kintex-7, Virtex-7, and Zynq-7000, each designed with different performance and cost trade-offs but all follow the core 7-Series architecture.

    Figure~\ref{fig:hierarchy_4} illustrates a high-level view of the hierarchical organization of a 7-Series FPGA. 
    At its lowest level, the device consists of a large array of atomic components called \emph{Basic Elements of Logic}~(\textbf{BELs}). 
    These BELs encompass look-up tables (\textbf{LUTs}), flip-flops (\textbf{FFs}), block RAMs (\textbf{BRAMs}), DSP slices (\textbf{DSPs}), and the configurable interconnect fabric. 
    They constitute the fundamental building blocks for implementing digital circuits on the FPGA.
    Note how the Tile arrangement is columnar, where each column will have only one Tile type.

    To manage this complexity, Xilinx organizes these BELs into incrementally abstract structures. 
    First, \textbf{BELs} are grouped into \textbf{Sites}. 
    Each Site is embedded into a \textbf{Tile}, and Tiles are further arranged into \textbf{Clock Regions}. 
    In some high-density devices, multiple Clock Regions may be consolidated into one or more \textbf{Super Logic Regions}~(SLRs). 
    However, for the scope of this paper, we focus on Xilinx 7-Series devices with only a single SLR.

    \end{multicols}
    {
        \centering
        \includegraphics[valign=c, width=0.3\columnwidth]{figures/hierarchy_5.png}
        \includegraphics[valign=c, width=0.6\columnwidth]{figures/hierarchy_4.png}
        \captionof{figure}{Architecture Hierarchy of a Xilinx FPGA}
        \label{fig:hierarchy_4}
    }
    \vfill
    \begin{multicols}{2}

    \newpage
    \subsection{CLB SLICEs}

    In the 7-Series architecture, the term \emph{CLB} (Configurable Logic Block) refers to a \emph{CLB Tile} that contains two \emph{SLICE} Sites. 
    Xilinx offers two variants of SLICE Sites: \textbf{SLICEL} and \textbf{SLICEM}.
    \begin{itemize}
        \item Each \texttt{SLICEL} has a set of BELs including eight \texttt{LUT}s, eight \texttt{FF}s, and one \texttt{CARRY4} adder. The \texttt{LUT} BELs in a \texttt{SLICEL} can only host \texttt{LUT} Cells. 
        \item The \texttt{SLICEM} includes all the features of a \texttt{SLICEL} but its \texttt{LUT} BELs can host both \texttt{LUT} Cells, which are asynchronous ROM elements, or \texttt{RAM32M} Cells, which are synchronous 32-deep RAM elements. These cells are also referred to as \emph{Distributed RAM} in the Xilinx documentation.
            These cells can offer an alternative to the larger, more dedicated 18K-36K \texttt{RAMB18E1} cells when RAM resources are highly utilized. 
    \end{itemize}
    In a typical 7-Series device, approximately 75\% of the SLICE Sites are \texttt{SLICEL}s and 25\% are \texttt{SLICEM}s. 
    A single CLB Tile can therefore host either two \texttt{SLICEL}s or one \texttt{SLICEL} and one \texttt{SLICEM}.
    To simplify the problem space, however, we will only consider \texttt{SLICEL}s for general logic and use the dedicated \texttt{RAMB18E1} cells to implement RAM elements.

    The BELs in these SLICEs facilitate the bulk of the general programmability of the FPGA fabric. 
    We will explain in detail the function and motivation behind these BELs.

    \textbf{LUTs} \quad
        Combinational logic is universal to all HDL designs. 
        As the their name suggests, a Look‑Up Table (LUT) map an input value to an output value. 
        LUTs facilitate combinational logic by acting as tiny asynchronously‑accessed ROMs whose contents are fixed when the FPGA is programmed.  
        For any boolean function, the synthesizer precalculates the boolean output to every possible input combination and stores the resulting truth table into a LUT's static memory. 
        The inputs are then essentially treated as an address space that maps to a data value space in an asynchronous ROM. 
        No explicit logic gates like NAND or XNOR are synthesized, contrary to what newcomers might expect from a "Field Programmable Gate Array".

        {
            \centering
            \includegraphics[width=\columnwidth]{figures/lut_synthesis.png}
            \captionof{figure}{LUT synthesis from user design}
            \label{fig:lut_synthesis}
        }
        \vspace{1.0cm}

        In the 7‑Series devices, one LUT can facilitate any 6-input boolean function, or two 5-input functions, as long as they share the same input signals.  
        The LUT can also host two independent boolean functions of up to 3 inputs each, even when the inputs are not shared.  
        Functions requiring more than six unique inputs are decomposed across multiple cascaded LUTs.
        Figure \ref{fig:lut_synthesis} shows an example of where a LUT is typically synthesized in a design entry.

    \textbf{FFs} \quad
        FFs are synthesized to facilitate synchronous event-driven signal assignment. 
        For most Verilog users, this generally means signal assignments wrapped in \texttt{always @(posedge clk)} statements. 
        Figure \ref{fig:ff_synthesis} shows an example of where a FF is typically synthesized. 
        The cell primitive \textbf{FDRE} is a type of FF and belongs to a family of D Flip Flops (DFFs) with Clock Enable (\texttt{CE}). 
        \begin{itemize}
            \item \texttt{FDCE} - DFF with CE and Asynchronous Clear
            \item \texttt{FDPE} - DFF with CE and Asynchronous Preset
            \item \texttt{FDSE} - DFF with CE and Synchronous Set
            \item \texttt{FDRE} - DFF with CE and Synchronous Reset
        \end{itemize}
        In a typical HDL design, the vast majority of FFs will be synthesized as FDREs with the occasional FDSE, as it is generally good practice to keep FPGA designs synchronous. 
        A FF BEL may also host a LATCH Cell, however, since they are generally bad practice in FPGA design, we will not consider latches in this paper.

        {
            \centering
            \includegraphics[width=\columnwidth]{figures/ff_synthesis.png}
            \captionof{figure}{FF synthesis from user design}
            \label{fig:ff_synthesis}
        }
        \vspace{1.0cm}

        Up to eight (8) FFs can be placed within the same SLICE, but only if they all share a common Clock-Enable (CE) net and a common Set-Reset (SR) net.
        This is because the SLICE has only one CE pin and one SR pin to interface with general routing. 
        The CE and SR signals from these pins are broadcast intra-Site to all FFs within.

    \newpage
    \textbf{LUT-FF Pairs} \quad
        FPGA designs are very often modelled as a collection of Finite State Machines (FSM) like shown in Figure \ref{fig:fsm}.
        Many times a design will also use pipelining, either to model signal buffers or shift registers, or to split up large combinational logic blocks into time slices to meet timing constraints. 
        These common design structures result in many consecutive sections of combinational logic feeding into a vector of registers. 
        The synthesizer naturally synthesizes these structures as consecutive pairs of LUTs feeding into FFs as shown in Figure \ref{fig:pipelining}. 
        Figure \ref{fig:lut_ff_pair} shows an example of a synthesized LUT-FF Pair.

        Shown in Figure \ref{fig:intersite_intrasite} are two possible placements for a LUT-FF Pair on the physical device. 
        On the right, the cells are placed across different Sites, thus the only way to route the net between the cells is through general inter-site routing. 
        On the bottom left, the cells are placed within the same Site in the same lane, taking advantage of the intra-site routing without burdening the general router with additional inter-site routing. 
        This is an important consideration to make while minimizing wirelength during placement.

        \vspace{1.0cm}
        {
            \centering
            \includegraphics[width=\columnwidth]{figures/fsm.png}
            \captionof{figure}{Finite state machine (Moore)}
            \label{fig:fsm}
        }
        \vspace{1.0cm}

        {
            \centering
            \includegraphics[width=0.9\columnwidth]{figures/pipelining.png}
            \captionof{figure}{Pipelining synthesized as consecutive LUT-FF pairs}
            \label{fig:pipelining}
        }

        \vfill
        .
        \newcolumn

        {
            \centering
            \includegraphics[width=0.9\columnwidth]{figures/lut_ff_pair.png}
            \captionof{figure}{A synthesized LUT-FF Pair}
            \label{fig:lut_ff_pair}
        }
        \vspace{1.0cm}

        {
            \centering
            \includegraphics[width=\columnwidth]{figures/intersite_intrasite_2.png}
            \captionof{figure}{Intrasite vs Intersite LUT-FF Placement}
            \label{fig:intersite_intrasite}
        }

    \newpage
    \textbf{CARRY} \quad
        An FPGA design will also typically implement many adders, counters, subtractors, or comparators, all of which are based on binary addition. 
        They are so ubiquitous that every that in the 7-Series architecture, every SLICE features a \texttt{CARRY4} BEL -- a 4-bit carry-lookahead (CLA) adder, a much better alternative to synthesizing adders via LUTs.

    \textbf{CARRY Chains} \quad
        These \texttt{CARRY4} blocks can be chained across SLICEs to implement wide adders efficiently. 
        The \texttt{CARRY4} BELs \emph{must} be chained vertically consecutively across SLICEs as the Carry-In (\texttt{CI}) and Carry-Out (\texttt{CO}) pins can only be routed this way. 
        A \texttt{CARRY4} cell may also directly connect to LUTs and FFs, and should be placed in the same Site whenever possible to minimize wirelength. 
        Shown in figure \ref{fig:carry_cell_group} shows how a \texttt{CARRY4} chain and associated LUTs and FFs can be placed across SLICEs.

        {
            \centering
            \includegraphics[width=\columnwidth]{figures/carry_cell_group.png}
            \captionof{figure}{A \texttt{SLICEL} with a \texttt{CARRY4} cell, 4 LUT cells, and 4 FF cells placed inside as shown in the Vivado device viewer}
            \label{fig:carry_cell_group}
        }

        {
            \centering
            \includegraphics[valign=c, width=4.0cm]{figures/carry_chain_placement.png}
            \includegraphics[valign=c, width=4.0cm]{figures/carry_chain_routes_3.png}
            \captionof{figure}{
                A \texttt{CARRY4} chain of size 3 placed across 3 SLICEs.
                \textbf{Left:} Simplified view, \textbf{Right:} As shown in the Vivado device viewer.
            }
            \label{fig:carry_cell_group}
        }

        \end{multicols}
        % \vfill
        {
            \centering
            \includegraphics[valign=c, width=3.0cm]{figures/adder_synthesis.png}
            \includegraphics[valign=c, width=14.0cm]{figures/carry_chain_edif_3.png}
            \captionof{figure}{A \texttt{CARRY4} chain of size 3 as shown in the Vivado netlist viewer}
            \label{fig:carry_chain_edif}
        }
        \begin{multicols}{2}
        \vspace{1.0cm}
        .

    \newpage
    \subsection{DSP Slices}

    \textbf{DSPs} \quad 
        FPGAs are often used as low latency Digital Signal Processing (DSP) accelerators. 
        Common DSP subsystems like Finite Impulse Response (FIR) filters, Fast Fourier Transform (FFTs), and convolutional neural nets (CNNs) demand fast large scale multiply-accumulate (MAC) capabilities. 
        The 7-Series arhictecture integrates DSP BELs into the logic fabric called DSP48E1 that can facilitate MAC efficiently. 
        The architecture hierarchy for DSPs is simple compared to CLBs and SLICEs. 
        A \texttt{DSP48E1} Tile contains two \texttt{DSP48E1} Sites, each containing one \texttt{DSP48E1} BEL, which can host a \texttt{DSP48E1} Cell.

        {
            \centering
            \includegraphics[width=\columnwidth]{figures/dsp_diagram.png}
            \captionof{figure}{
                Basic \texttt{DSP48E1} Slice Functionality
            }
            \label{dsp_diagram}
        }
        \vspace{0.5cm}

    \textbf{DSP Cascades} \quad 
        Higher level DSP functions are supported by cascading \texttt{DSP48E1} slices in a \texttt{DSP48E1} column.
        Much like \texttt{CARRY4} chains, \texttt{DSP48E1} cascades must necessarily be placed vertically consecutively across \texttt{DSP48E1} Sites. 
        They are connected by three busses: \texttt{ACOUT} to \texttt{ACIN}, \texttt{BCOUT} to \texttt{BCIN}, \texttt{PCOUT} to \texttt{PCIN}.
        These signal busses run directly between the vertical \texttt{DSP48E1} slices without burdening the general routing resources. 
        The ability to cascade this way provides a high-performance and low-power implementation of digital signal processing systems.

        \vspace{1.0cm}
        {
            \centering
            \includegraphics[valign=c, width=2.5cm]{figures/dsp_cascade_placement.png}
            \includegraphics[valign=c, width=3.0cm]{figures/dsp_cascade_routed.png}
            \captionof{figure}{
                A \texttt{DSP48E1} cascade of size 2 placed across 2 \texttt{DSP48E1} Sites.
                \textbf{Left:} Simplified view, \textbf{Right:} As shown in the Vivado device viewer.
            }
        }

        \end{multicols}
        {
            \centering
            \includegraphics[valign=c, width=4.0cm]{figures/mult_synthesis.png}
            \includegraphics[valign=c, width=13.0cm]{figures/dsp_synthesis.png}
            \captionof{figure}{
                Simple multiplier synthesis from user design.
            }
            \label{mult_synthesis}
        }
        \begin{multicols}{2}

    \newpage

    \subsection{Block RAM}
        In addition to SLICEs and DSPs, the 7-Series also offers dedicated Block Random Access Memory (BRAM) BELs. 
        These BRAMs come in two variants: \textbf{RAMB18E1} and \textbf{RAMB36E1}. 
        \begin{itemize}
            \item \texttt{RAMB18E1} - 
            Has a capacity of 18 Kilobits. 
            It can be configured as single port RAM with dimensions ranging between (1-bit wide by 16K deep) to (18-bit wide by 1024 deep). 
            It can also be configured as a (36-bit wide by 512 deep) true simple dual port RAM. 
            \item \texttt{RAMB36E1} - 
            Has a capacity of 36 Kilobits. 
            It can be configured as single port RAM with dimensions ranging between (1-bit wide by 32K deep) to (36-bit wide by 1024 deep). 
            It can also be configured as a (72-bit wide by 512 deep) simple dual port RAM. 
        \end{itemize}

        One BRAM Tile contains one \texttt{RAMB36E1} Site and two \texttt{RAMB18E1} Sites. 
        The \texttt{RAMB36E1} Site contains one \texttt{RAMB36E1} BEL, which can host one \texttt{RAMB36E1} Cell. 
        Likewise, the \texttt{RAMB18E1} Site contains one \texttt{RAMB18E1} BEL, which can host one \texttt{RAMB18E1} Cell.

        Like DSP cascades, BRAMs may also be cascaded together in a column to implement large memories efficiently, with some intermediate signals between them routed intra-Tile without burdening the general routing. 
        However, unlike DSPs, large memories decomposed amongst multiple BRAMs can also be routed together through general routing. 
        Furthermore, in most design scenarios, large memories will not utilize the intra-Tile signals.

        To simplify the problem space, we will not constrain large BRAMs to be cascaded together in consecutive vertical Tiles. 
        We can essentially treat \texttt{RAMB18E1} and \texttt{RAMB36E1} Cells as loose, minimally constrained single cells, in contrast to the highly constrained \texttt{DSP48E1} cascades and \texttt{CARRY4} chains.

        {
            \centering
            \includegraphics[valign=c, width=3.0cm]{figures/bram.png}
            \includegraphics[valign=c, width=2.0cm]{figures/bram_tile.png}
            \captionof{figure}{
                A BRAM Tile containing two \texttt{RAMB18E1} Sites and one \texttt{RAMB36E1} Site.
                \textbf{Left:} simplified view. \textbf{Right:} as seen in the device viewer, BELs highlighted in white. 
            }
            \label{fig:bram_tile}
        }

        {
            \centering
            \includegraphics[valign=c, width=4.0cm]{figures/bram_synthesis.png}
            \includegraphics[valign=c, width=\columnwidth]{figures/bram_edif.png}
            \captionof{figure}{
                An example of BRAM synthesis (via inference)
            }
            \label{fig:bram_synthesis}
        }

    \subsection{Further Documentation}

    For more in-depth details about 7-Series FPGAs, refer to the official Xilinx user guides such as:
    \begin{itemize}
        \item \emph{7 Series FPGAs Overview} (UG476)
        \item \emph{7 Series FPGA Configurable Logic Block} (UG474)
        \item \emph{7 Series Memory Resources} (UG473)
        \item \emph{7 Series DSP48E1 Slice} (UG479)
    \end{itemize}

    This architectural context provides the necessary background for understanding how a placement algorithm should account for resource constraints and optimize performance in modern FPGA designs.

\newpage
\section{FPGA Design Flow and Toolchain}
    \label{sec:fpga_flow_toolchain}

    Modern FPGA designs require a sophisticated toolchain to bridge the gap between high-level hardware descriptions and the final bitstream used to configure the FPGA. 
    Figure~\ref{fig:design_flow} illustrates a representative process that converts an abstract Hardware Description Language (HDL) design into a verified configuration file for a target device.

    {
        \centering
        \includegraphics[width=0.9\columnwidth]{figures/design_flow.png}
        \captionof{figure}{A typical FPGA design and verification workflow.}
        \label{fig:design_flow}
    }

    \begin{enumerate}
        \item \textbf{Design Entry:} 
            An engineer describes the intended functionality of the digital system using a hardware description language (HDL) such as Verilog or VHDL. 
            During this phase, the coding style can vary (behavioral, structural, dataflow, etc.), but it always aims to capture high-level behavior rather than device-specific details.

        \item \textbf{Synthesis:} 
            The synthesis tool parses the HDL source, performs logical optimizations, and maps the design onto primitive cells that suit the target FPGA technology. 
            The output is typically a structural \texttt{netlist} (e.g., EDIF or structural Verilog) which details how the design’s logic is broken down into LUTs, FFs, and other vendor-specific cells.

        \item \textbf{Placement and Routing (Implementation):} 
            In \emph{placement}, each logical cell from the synthesized netlist is assigned to a physical location on the FPGA fabric. 
            For instance, LUTs and FFs go into specific \emph{BELs} within the device’s CLB sites, and specialized cells such as DSPs and Block RAMs must be placed in their corresponding tile types. 
            Next, \emph{routing} determines how signals are physically wired through the FPGA’s configurable interconnect network. 
            Modern tools often interleave these steps (e.g., fluid-placement routing or routing-aware placement) to better meet timing and area objectives.

        \item \textbf{Bitstream Generation:} 
            After a design is fully placed, routed, and timing-closed, the toolchain produces a final \emph{bitstream} that sets the configuration of every programmable element in the FPGA. 
            This bitstream can then be loaded onto the device, either through vendor software or via a custom programming interface.

        \item \textbf{Verification:} 
            In parallel to the design flow, simulations and testbenches validate correctness of the user's design at multiple abstraction levels. 
            Engineers may begin with behavioral simulations, then progress to post-synthesis simulations, and finally to post-implementation simulations that incorporate estimated routing delays. 
            With each higher level of fidelity, computational requirements grow significantly due to increasing complexity and the need to analyze more variables over time. 
            Ensuring correct functionality and meeting timing closure at the post-implementation stage is crucial before deploying the design to hardware. 
            Given the importance of thorough verification, many established companies dedicate one verification engineer for every design engineer.

    \end{enumerate}

    % This workflow underscores the critical role of \textbf{placement} in bridging the netlist to a physical realization. 
    % An efficient placement algorithm can drastically reduce compile times and improve design performance, enabling broader adoption of FPGAs in application spaces that require fast design iteration.

\newpage
\section{RapidWright API}
    \label{sec:rapidwright_api}

    \textbf{RapidWright} is an open-source Java framework from AMD/Xilinx that provides direct access to the netlist and device databases used by vendor tools. 
    This framework positions itself as an additional workflow column, allowing users to intercept or replace stages of the standard design flow with custom optimization stages (see Figure~\ref{fig:vivado_dcps}).

    \begin{itemize}
    \item \textbf{Design Checkpoints:} 
        RapidWright leverages \texttt{.dcp} files (design checkpoints) generated at various stages of a Vivado flow. 
        By importing a checkpoint, engineers can manipulate the netlist, placement, or routing externally, then re-export a modified checkpoint for further processing in the Vivado workflow column.

    \item \textbf{Key Packages:} 
        RapidWright revolves around three primary data model packages:
        \begin{enumerate}
        \item \emph{edif} -- Represents the logical netlist in an abstracted EDIF-like structure.
        \item \emph{design} -- Contains data structures for the physical implementation (Cells, Nets, Sites, BELs, etc.).
        \item \emph{device} -- Provides a database of the target FPGA architecture (e.g., Site coordinates, Tile definitions, routing resources).
        \end{enumerate}

    \item \textbf{Interfacing with the Netlist and Device:} 
        An engineer can query the netlist to find specific resources (LUTs, FFs, DSPs, etc.) and then map or move them onto device sites. 
        This level of control over backend resources is necessary for research in custom placement, advanced packing techniques, or experimental routing algorithms.
    \end{itemize}

    {
        \centering
        \includegraphics[width=0.8\columnwidth]{figures/vivado_dcps.png}
        \captionof{figure}{RapidWright workflow integrating into the default Vivado design flow.}
        \label{fig:vivado_dcps}
    }

    By exposing these low-level internals, RapidWright allows fine-grained design transformations that go beyond the standard Vivado IDE’s capabilities. 
    Researchers can prototype new EDA strategies without needing to re-implement an entire FPGA backend from scratch, thus accelerating innovation in placement and routing methodologies.

\newpage
\section{What is a Netlist?}
    \label{sec:netlist}
    In its simplest form, a netlist is a list of every component in an electronic design and a list of nets they connect to. 
    Generally, a net, or network, denotes any group of two or more interconnected elements - whether they be transistors, logic gates, or in our case, logical cells. 
    More precisely, nets attach to component pins, and those pins reside on cells, each of which exposing multiple ports. 
    In this way, a netlist makes explicit both the individual building blocks of a circuit and the precise wiring that links them.

    In Vivado, a Netlist can be synthesized as a Hierarchical or a Flattened netlist. 
    Figure ~\ref{fig:hierarchical_design} shows the Verilog code for a simple HDL design using module instance hierarchy. 
    Figure ~\ref{fig:hier_netlist} shows the design synthesized into a hierarchical netlist. 
    The synthesizer attempts to construct the module hierarchy as close to the module instantiation hierarchy as defined by the raw user design entry. 
    Figure ~\ref{fig:flat_netlist} shows the same design synthesized into a flat netlist. 
    In a synthesized Netlist, the leaf cells (lowest-level cells) are necessarily primitive cells from the architecture primitives library which includes \texttt{LUT6}, \texttt{FDRE}, \texttt{CARRY4}, \texttt{DSP48E1}, etc..

    \end{multicols}
    {
        \raggedright
        \includegraphics[valign=t, scale=0.3]{figures/netlist_synth/top_level.png}
        \includegraphics[valign=t, scale=0.3]{figures/netlist_synth/module_0.png}
        \includegraphics[valign=t, scale=0.3]{figures/netlist_synth/module_1.png}
        \includegraphics[valign=t, scale=0.3]{figures/netlist_synth/module_2.png}
        \includegraphics[valign=t, scale=0.3]{figures/netlist_synth/module_3.png}
        \captionof{figure}{A simple HDL design with module hierarchy.}
        \label{fig:hierarchical_design}
    }
    {
        \centering
        \includegraphics[valign=c, width=11.5cm]{figures/netlist_synth/hier_netlist.png}
        \includegraphics[valign=c, width=6cm]{figures/netlist_synth/hier_graph.png}
        \captionof{figure}{
            \textbf{Left:} A hierarchical netlist consisting of LUTs and FFs.
            \textbf{Right:} The cell hierarchy graph.
        }
        \label{fig:hier_netlist}
    }
    \vspace{0.5cm}
    {
        \centering
        \includegraphics[valign=c, width=10cm]{figures/netlist_synth/flat_netlist.png}
        \includegraphics[valign=c, width=4cm]{figures/netlist_synth/flat_graph.png}
        \captionof{figure}{
            \textbf{Left:} A flattened netlist consisting of LUTs and FFs.
            \textbf{Right:} The flattened cell hierarchy graph.
        }
        \label{fig:flat_netlist}
    }
    \begin{multicols}{2}

    \subsection{Netlist Manipulation in RapidWright}

    One of RapidWright's most powerful features is netlist manipulation via its \texttt{edif} package. 
    A netlist can first be extracted from a \texttt{.dcp} design checkpoint like the following:

    SHOULD I INCLUDE A RAPIDWRIGHT HELLO WORLD EXAMPLE JUST TO GIVE AN IMPRESSION?

    OR CAN I JUST DIRECT THE READER TO THE RAPIDWRIGHT DOCUMENTATION?

    \end{multicols}
    {
        \centering
        \includegraphics[width=\columnwidth]{figures/raw_edif.png}
        \captionof{figure}{A raw EDIF Netlist with uncountably many cells}
        \label{fig:raw_edif}
    }
    \begin{multicols}{2}

    .

\newpage
\section{Simulated Annealing}
    \label{sec:simulated_annealing}
    With a basic understanding of FPGA architecture, design placement, and RapidWright, we have all the necessary pieces to implement our SA placer. 
    Here we outline in detail each substage of our implementation: PrePacking, Packing, and Placement. 
    Shown in Figure \ref{fig:edif_design_device} shows an overview of the placement workflow.

    {
        \centering
        \includegraphics[width=0.9\columnwidth]{figures/edif_design_device.png}
        \captionof{figure}{Our placement workflow}
        \label{fig:edif_design_device}
    }

    \end{multicols}
    {
        \centering
        \includegraphics[width=0.8\columnwidth]{figures/substages.png}
        \captionof{figure}{The data classes populated at each substage: PrepackedDesign, PackedDesign, and PlacedDesign.}
        \label{fig:substages}
    }
    \begin{multicols}{2}

    \subsection{Prepacking}
        \label{subsec:prepacking}

        The first step in our placement flow is \textbf{prepacking}. 
        Recall from the 7-Series architecture that there are certain cell structures that must adhere to certain placements constraints to ensure legality, and by design, to minimize wirelength. 
        The job of the prepacker is to traverse the raw EDIF netlist, detect these cell structures, and consolidate these cells into clusters or groups of clusters that naturally reflect these placement constraints.

>>>>

\subsection{Site Packing}
\label{subsec:packing}
